---
title: "Moment-Generating Function based Infill-criterion for Bayesian Optimization"
author: 'Hao Wang'
output: 
  html_notebook:
    highlight: zenburn
    theme: spacelab
    toc: true
    toc_float: true
    number_sections: true
date: '2017-06-13'
---
\newcommand{\E}{\mathbb{E}}

```{r setup, echo=FALSE}
# setup for all the R chunks
knitr::opts_knit$set(width = 1000)
# change the workind dir for all the chunks
knitr::opts_knit$set(root.dir = '~/Dropbox/code_base/InfillCriteria/R/') 
```
```{r, include=FALSE}
library(reshape2)
library(ggplot2)
library(dplyr)
library(latex2exp)
library(DiceOptim)
library(DiceKriging)
library(extrafont)
library(tikzDevice)
loadfonts()   # exposing the fonts to plotting library

# check CM Roman font. Install it if does not exist
if (!'CM Roman' %in% fonts()) {
  font_install('fontcm', prompt = F)
  loadfonts()
}

source('./criteria.R')
source('./fitness.R')
theme.my <- theme_grey() + 
  theme(text = element_text(size = 30, family = 'Latin Modern Math'),
        legend.key.size = unit(2, "line"),
        plot.title = element_text(hjust = 0.5))
theme_set(theme.my)
```

# Motivation

How to characterise the distribution of improvements? (higer) Moments? Given its existence, moment-generating function (MGF) could be a natural choice as it combines all (infinite) moments together:
$$M_{X}(t) = \E[e^{tX}] = 1 + t\E X + \frac{t^2\E X^2}{2!} + \frac{t^3\E X^3}{3!} + \cdots$$
Plot the MGF function when varying the $t$ value (as below). Judging from the plot, the $t$ value should be scaled log-linearly as higher $t$ values yield quite similar MGF functions. The similar phenomemone can be observed from Generalized Expected Improvement.

# Illustration on 1D Ackley function 
We create a example GPR on the well-known Ackley function:
```{r}
set.seed(11)
lower <- -5
upper <- 7

f <- . %>% {sin(.)}
f <- ackley
doe <- runif(6, lower, upper) %>% data.frame(x = .)
response <- sapply(doe$x, f)

gp <- km(~1, doe, response, covtype = 'matern3_2',
         nugget = 1e-10,
         control = list(trace = F,
                        pgtol = 1e-15, 
                        factr = 1e4,
                        maxit = 1e5))

x <- c(seq(lower, upper, length.out = 400), doe$x)
res <- predict(gp, x, 'UK', checkNames = F) 
df <- data.frame(x, y = sapply(x, f), y.hat = res$mean, se = res$sd)
```

## MGF-based 
Preparation of the MGF-based Acquisition function data.
```{r}
df.MGF <- data.frame(x)
# t value should be scaled log-linearly to prevent similar MGF functions when t is big
for (t in exp(seq(-2, log(5), length.out = 30))) {
  df.MGF %<>% mutate(tmp = sapply(x, . %>% MGF(model = gp, t = t))) %>%
    mutate(tmp = 10 * tmp / max(tmp)) %>%
    rename_(.dots = setNames(list('tmp'), as.character(t)))
}

df.MGF %<>% melt(id = 'x', variable.name = 't', value.name = 'MGF') %>%
  mutate(t = as.numeric(levels(t))[t]) 

df.MGF.max <- df.MGF %>% 
  group_by(t) %>% 
  summarise(x_max = x[MGF == max(MGF)], max = max(MGF))
```
And then plot the normalized (for the better visualization) acquisition functions:
```{r, fig.height=8, fig.width=12, fig.cap=' '}
tikz(file = "MGFI.tex", width = 14, height = 9)
p <- ggplot(df.MGF) + 
  geom_line(aes(x, MGF, colour = t, group = t), size = 1, alpha = .8) + 
  geom_ribbon(data = df, aes(x = x, ymin = y.hat - 1.96 * se, ymax = y.hat + 1.96 * se, fill = '95\\% CI'), alpha = 0.19) +
  geom_line(data = df, aes(x, y, linetype = 'function'),  size = 1, alpha = 0.5) +
  geom_line(data = df, aes(x, y.hat, linetype = 'prediction'), size = 1, alpha = 0.5) +
  geom_point(data = data.frame(x = doe$x, y = response), aes(x, y),
             colour = 'black', alpha = .8, shape = 20, size = 6) +
  geom_point(data = df.MGF.max, aes(x_max, max), colour = 'black', alpha = .9, shape = 3, size = 10) +
  scale_colour_gradientn(name = '$\\mathit{t}$', breaks = quantile(unique(df.MGF$t)), 
                         colours = colorspace::rainbow_hcl(10), trans = "log", 
                         labels = function(x) sprintf("%.2f", x)) +
  scale_linetype_manual(name = NULL, values = c('prediction' = 'solid', 'function' = 'dashed'), guide = FALSE) + 
  scale_fill_manual(name = NULL, values = c('95\\% CI' = '#26989C')) + 
  labs(x = '$\\mathit{x}$', y = '$\\mathit{f}$') 

print(p)
dev.off()
# ggsave('MGF.pdf', plot = p, width = 14, height = 9, device = cairo_pdf)
# embed_fonts('MGF.pdf', outfile = 'MGF2.pdf')
```
The infinite weights of all the moments are plotted as below, which is actually the pmf of Poisson distribution.
```{r, fig.height=7, fig.width=10}
w <- function(t, n) {
  x <- seq(0, n - 1)
  weights <- t ^ x / factorial(x)
  return(weights / exp(t))
}

N <- 40
t <- c(0.5, seq(1, 21, length.out = 5))
weights <- sapply(t, . %>% w(N))
colnames(weights) <- t

df <- cbind(seq(0, N - 1), weights) %>% 
  as.data.frame %>%
  melt(id.vars = 'V1', variable.name = 't', value.name = 'weight')

p <- ggplot(df, aes(x = V1, y = weight, colour = t)) + 
  geom_point(size = 3) + geom_line(size = 1) + 
  scale_color_discrete(name = TeX('$\\mathit{t}$')) +
  labs(x = 'order of the moments')

ggsave('_weight.pdf', plot = p, width = 14, height = 9)
embed_fonts('_weight.pdf', outfile = 'weight.pdf')
```

## Generalized EI
This MGF-based infill-criterion is underpinned by the similar rationale behind the so-called generalized expected Improvement (GEI). Those two funtions are compared in the following.
Generate some GEI functions by varying the $g$ value:
```{r}
df.GEI <- data.frame(x)
for (g in seq(0, 10)) {
  df.GEI %<>% mutate(tmp = sapply(x, . %>% GEI(model = gp, g = g))) %>%
    mutate(tmp = 10 * tmp / max(tmp)) %>%
    rename_(.dots = setNames(list('tmp'), as.character(g)))
}

df.GEI %<>% melt(id = 'x', variable.name = 'g', value.name = 'GEI') 
df.GEI.max <- df.GEI %>% 
  group_by(g) %>% 
  summarise(x_max = x[GEI == max(GEI)], max = max(GEI))
```

```{r, fig.height=8, fig.width=12}
tikz(file = "GEI.tex", width = 14, height = 9)
p <- ggplot(df.GEI) + 
  geom_line(aes(x, GEI, colour = g, group = g), size = 1, alpha = 0.8) + 
  geom_ribbon(data = df, aes(x = x, ymin = y.hat - 1.96 * se, ymax = y.hat + 1.96 * se, fill = '95\\% CI'), alpha = 0.19) +
  geom_line(data = df, aes(x, y, linetype = 'function'),  size = 1, alpha = 0.5) +
  geom_line(data = df, aes(x, y.hat, linetype = 'prediction'), size = 1, alpha = 0.5) +
  geom_point(data = data.frame(x = doe$x, y = response), aes(x, y), colour = 'black', alpha = .8, shape = 20, size = 6) +
  geom_point(data = df.GEI.max, aes(x_max, max), colour = 'black', alpha = .9, shape = 3, size = 10) +
  scale_color_discrete(name = '$\\mathit{g}$') +
  scale_linetype_manual(name = NULL, values = c('prediction' = 'solid', 'function' = 'dashed'), guide = FALSE) + 
  scale_fill_manual(name = NULL, values = c('95\\% CI' = '#26989C')) + 
  labs(x = '$\\mathit{x}$', y = '$\\mathit{f}$')

print(p)
dev.off()
# ggsave('_GEI.pdf', plot = p, width = 14, height = 9)
# embed_fonts("_GEI.pdf", outfile = "GEI.pdf")
```